Potential and limits of site-specific substitution model inference.

### 2020-01-13 manuscript/ms.tex
fixed typos

\begin{abstract}
Natural selection imposes a complex filter on which variants persist in a population resulting in evolutionary patterns that vary greatly along the genome.
Some sites evolve close to neutral, while others are highly conserved, allow only specific states, or only change in concert with other sites.
Most commonly used evolutionary models, however, ignore much of this complexity and at best account for variation in the rate at which different sites change.
Here, we present an efficient algorithm to estimate more complex models that allow for site-specific preferences and explore the accuracy at which such models can be estimated from simulated data.
We find that the interative approximate maximum likelihood schemes uses information in the data efficiently and results in accurate  estimates of site specific preferences from large data sets with moderately diverged sequences.
We find that ignoring site specific preferences during branch length estimation of phylogenetic trees -- an assumption of all popular phylogeny software -- results in substantial underestimation comparable to the error incurred when ignoring rate variation.
However, the joint estimation of branch lengths, site-specific rates, and site-specific preferences can suffer from identifiability problems is typically unable to recover the correct branch lengths.
Site-specific preferences estimated from large HIV \emph{pol} alignments show good concordance with intra-host estimates and highlight the inability of estimating deep divergence times of HIV and SIVs using molecular clock approaches due to saturation.
\end{abstract}

@@

These conditions can be solved iteratively. We derive these update rule here $\mu^b$ and refer to the appendix for the other (more laborious) update rules.
\begin{equation}
	\frac{\partial \ell(\{\vec{s}\}|T,\mat{Q})}{\partial \mu^b} = \sum_{k\in T} \frac{t_k}{\mu^b}\frac{\sum_i Q^b_{k^b_c,i}\left( e^{Q^bt_k} \right)_{i, k^b_p}}{\left( e^{Q^bt_k} \right)_{k^b_c, k^b_p}}
\end{equation}
The individual terms in this sum behave very differently for cases where $k^b_c=k^b_p$ (no change at site $b$ on branch $k$) and $k^b_c\neq k^b_p$ (at least one mutation).
In the limit of short branches $\mu^b t_k\ll 1$, we can expand the matrix exponential $e^{\mat{Q}t}=\delta_{ij} + \mat{Q}t+\cdots$ to obtain more insightful expressions and we will separate branches with $k^b_c=k^b_p$  and $k^b_c\neq k^b_p$ right away
\begin{equation}
\begin{split}
	\frac{\partial \ell(\{\vec{s}\}|T,\mat{Q})}{\partial \mu^b} &\approx \sum_{k\in T, k^b_c=k^b_p} \frac{t_k}{\mu^b}\frac{\sum_i Q_{k^b,i}(\delta_{i,k^b} + t_k Q^b_{i,k^b}\cdots)}{1 - t_k\mu^b\sum_{j}\eqp_j W_{k^bj}+\ldots} +
	\sum_{k\in T, k^b_c\neq k^b_p} \frac{t_k}{\mu^b}\frac{\sum_i Q_{k^b_c,i}(\delta_{i,k^b_p} + t_k Q^b_{i,k^b_p}\cdots)}{t_k\mu^b\eqp_{k^b_c} W_{k^b_ck^b_p}+\ldots} \\
		&\approx -\sum_{k\in T, k^b_c=k^b_p}t_k \sum_i \eqp_{i}W_{i,k^b} +
	\sum_{k\in T, k^b_c\neq k^b_p} \frac{t_k}{\mu^b}\frac{p_{k^b_c}W_{k^b_c, l^b_p} + \cdots}{t_k\eqp_{k^b_c} W_{k^b_ck^b_p}+\ldots} \\
		&\approx -\sum_{k\in T, k^b_c=k^b_p}t_k \sum_i \eqp_{i}W_{i,k^b} + 	\sum_{k\in T, k^b_c\neq k^b_p} \frac{1}{\mu^b} \\
		&= -\sum_{ij}\eqp_{i}W_{i,j}\tau_j^b + \frac{1}{\mu^b}\sum_{ij} n^b_{ij}
	\end{split}
\end{equation}
where $\tau^b_j$ is the sum of all branch length along which site $b$ is in state $j$ and $n^b_{ij}$ is the number of times the sequences at site $b$ changes from $j$ to $i$ along branches of the tree.
This condition (and the corresponding ones in the supplement suggest solution at fixed $\tau^b_j$ and $n^b_{ij}$ using the following iterative update:
\begin{equation}
\label{eq:update_app}
	\begin{split}
		\mu^b & \leftarrow \mu^b\frac{\pc+\sum_{ij} n^b_{ij}}{\sum_{ij}\mu^b \eqp_{i}W_{ij}\tau_j^b} \\
		\eqp^b_i & \leftarrow \eqp^b_i\frac{\pc+\delta_{is^{b}} + \sum_{j\neq i} n^b_{ij}}{\eqp^b_i(q\pc+1) + \mu^b\eqp^b_i \sum_{j}W_{ij}\tau_j^b} \\
		W_{ij} & \leftarrow W_{ij}\frac{\sum_b (n^b_{ij}+n^b_{ji})}{\sum_b \mu^b W_{ij}(\eqp_i^b\tau_j^b + \eqp_j^b \tau_i^b)} \\
	\end{split}
\end{equation}
where $\pc$ is a pseudocount analogous to a Dirichlet prior that in the absence of data will drive the $\eqp_i^b$ to a flat distribution and the substitution rates to $\pc$ over the total tree length.
To make the crucial features of these update rules more explicit, we have multiplied numerator and denominator with the parameter that is being updated. This (i) makes the update explicitly multiplicative and therefore ensures positivity, and (ii) illustrates that each of these rules are the ratio of the {\it observed} number of transitions between states $n^b_{ij}$ and the {\it expected} number $\eqp^b_i W_{ij}\tau_j^b$ -- each appropriately summed over sites or states.

>>>>

\begin{abstract}
Natural selection imposes a complex filter on which variants persist in a population resulting in evolutionary patterns that vary greatly along the genome.
Some sites evolve close to neutral, while others are highly conserved, allow only specific states, or only change in concert with other sites.
Most commonly used evolutionary models, however, ignore much of this complexity and at best account for variation in the rate at which different sites change.
Here, we present an efficient algorithm to estimate more complex models that allow for site-specific preferences and explore the accuracy at which such models can be estimated from simulated data.
We find that the interative approximate maximum likelihood scheme uses information in the data efficiently and results in accurate  estimates of site specific preferences from large data sets with moderately diverged sequences.
We find that ignoring site specific preferences during branch length estimation of phylogenetic trees -- an assumption of all popular phylogeny software -- results in substantial underestimation comparable to the error incurred when ignoring rate variation.
However, the joint estimation of branch lengths, site-specific rates, and site-specific preferences can suffer from identifiability problems and is typically unable to recover the correct branch lengths.
Site-specific preferences estimated from large HIV \emph{pol} alignments show good concordance with intra-host estimates and highlight the inability of estimating deep divergence times of HIV and SIVs using molecular clock approaches due to saturation.
\end{abstract}

@@

These conditions can be solved iteratively. We derive these update rule here $\mu^b$ and refer to the Materials and Methods for the other (more laborious) update rules.
\begin{equation}
	\frac{\partial \ell(\{\vec{s}\}|T,\mat{Q})}{\partial \mu^b} = \sum_{k\in T} \frac{t_k}{\mu^b}\frac{\sum_i Q^b_{k^b_c,i}\left( e^{Q^bt_k} \right)_{i, k^b_p}}{\left( e^{Q^bt_k} \right)_{k^b_c, k^b_p}}
\end{equation}
The individual terms in this sum behave very differently for cases where $k^b_c=k^b_p$ (no change at site $b$ on branch $k$) and $k^b_c\neq k^b_p$ (at least one mutation).
In the limit of short branches $\mu^b t_k\ll 1$, we can expand the matrix exponential $e^{\mat{Q}t}=\delta_{ij} + \mat{Q}t+\cdots$ to obtain more insightful expressions and we will separate branches with $k^b_c=k^b_p$  and $k^b_c\neq k^b_p$ right away
\begin{equation}
\begin{split}
	\frac{\partial \ell(\{\vec{s}\}|T,\mat{Q})}{\partial \mu^b} &\approx \sum_{k\in T, k^b_c=k^b_p} \frac{t_k}{\mu^b}\frac{\sum_i Q_{k^b,i}(\delta_{i,k^b} + t_k Q^b_{i,k^b}\cdots)}{1 - t_k\mu^b\sum_{j}\eqp_j W_{k^bj}+\ldots} +
	\sum_{k\in T, k^b_c\neq k^b_p} \frac{t_k}{\mu^b}\frac{\sum_i Q_{k^b_c,i}(\delta_{i,k^b_p} + t_k Q^b_{i,k^b_p}\cdots)}{t_k\mu^b\eqp_{k^b_c} W_{k^b_ck^b_p}+\ldots} \\
		&\approx -\sum_{k\in T, k^b_c=k^b_p}t_k \sum_i \eqp_{i}W_{i,k^b} +
	\sum_{k\in T, k^b_c\neq k^b_p} \frac{t_k}{\mu^b}\frac{p_{k^b_c}W_{k^b_c, l^b_p} + \cdots}{t_k\eqp_{k^b_c} W_{k^b_ck^b_p}+\ldots} \\
		&\approx -\sum_{k\in T, k^b_c=k^b_p}t_k \sum_i \eqp_{i}W_{i,k^b} + 	\sum_{k\in T, k^b_c\neq k^b_p} \frac{1}{\mu^b} \\
		&= -\sum_{ij}\eqp_{i}W_{i,j}\tau_j^b + \frac{1}{\mu^b}\sum_{ij} n^b_{ij}
	\end{split}
\end{equation}
where $\tau^b_j$ is the sum of all branch length along which site $b$ is in state $j$ and $n^b_{ij}$ is the number of times the sequences at site $b$ changes from $j$ to $i$ along branches of the tree.
This condition (and the corresponding ones in the supplement suggest solution at fixed $\tau^b_j$ and $n^b_{ij}$ using the following iterative update:
\begin{equation}
\label{eq:update_app}
	\begin{split}
		\mu^b & \leftarrow \mu^b\frac{\pc+\sum_{ij} n^b_{ij}}{\sum_{ij}\mu^b \eqp_{i}W_{ij}\tau_j^b} \\
		\eqp^b_i & \leftarrow \eqp^b_i\frac{\pc+\delta_{is^{b}} + \sum_{j\neq i} n^b_{ij}}{\eqp^b_i(q\pc+1) + \mu^b\eqp^b_i \sum_{j}W_{ij}\tau_j^b} \\
		W_{ij} & \leftarrow W_{ij}\frac{\sum_b (n^b_{ij}+n^b_{ji})}{\sum_b \mu^b W_{ij}(\eqp_i^b\tau_j^b + \eqp_j^b \tau_i^b)} \\
	\end{split}
\end{equation}
where $\pc$ is a pseudocount analogous to a Dirichlet prior that in the absence of data will drive the $\eqp_i^b$ to a flat distribution and the substitution rates to $\pc$ over the total tree length.
To make the crucial features of these update rules more explicit, we have multiplied numerator and denominator with the parameter that is being updated. This (i) makes the update explicitly multiplicative and therefore ensures positivity, and (ii) illustrates that each of these rules are the ratio of the {\it observed} number of transitions between states $n^b_{ij}$ and the {\it expected} number $\eqp^b_i W_{ij}\tau_j^b$ -- each appropriately summed over sites or states.
